Bring in first dataset (conumption)

```{r}
library(readxl)
temp = tempfile(fileext = ".xlsx") # use always with Excel
dataURL <- "https://github.com/ihsun-uw/Assignment-1/raw/master/consumptionnps4.xlsx" # link to data
download.file(dataURL, destfile=temp, mode='wb')  # file will be downloaded temporarily

cons = read_excel(temp, sheet =1)
cons=as.data.frame(cons)
```


```{r}
names(cons)
```

keeping necessary variables
```{r}
cons=cons[,c("y4_hhid", "expmR", "region")]
```

Second dataset (education)

```{r}
library(readxl)
temp = tempfile(fileext = ".xlsx") # use always with Excel
dataURL <- "https://github.com/ihsun-uw/Assignment-1/raw/master/Education.xlsx" # link to data
download.file(dataURL, destfile=temp, mode='wb')  # file will be downloaded temporarily

educ = read_excel(temp, sheet =1)
educ=as.data.frame(educ)
```

```{r}
names(educ)
```


```{r}
educ=educ[,c("y4_hhid", "indidy4", "hh_c02", "hh_c03", "hh_c07")]
```

keeping only the head of household information

```{r}
educ=educ[educ$indidy4==1,]
```

Third dataset with head of household gender information

```{r}

library(readxl)
temp = tempfile(fileext = ".xlsx") # use always with Excel
dataURL <- "https://github.com/ihsun-uw/Assignment-1/raw/master/individual.xlsx" # link to data
download.file(dataURL, destfile=temp, mode='wb')  # file will be downloaded temporarily

indiv = read_excel(temp, sheet =1)
indiv=as.data.frame(indiv)

```

```{r}
names(indiv)
```

```{r}
indiv=indiv[,c("y4_hhid", "indidy4", "hh_b02", "hh_b04")]
```

keeping only the head of household information
```{r}
indiv=indiv[indiv$indidy4==1,]
```

merge data
```{r}
merge1=merge(cons, educ)

alldata=merge(merge1, indiv)
```

```{r}
str(alldata)
```



Explore main variable of interest
```{r}
summary(alldata$'expmR')
```

```{r}
sd(alldata$'expmR',na.rm = T)
```

```{r}
data=alldata[is.finite(alldata$'expmR'),]

var=alldata$'expmR'

mnVar=mean(var,na.rm = T)
sdVar=sd(var,na.rm = T)

library(ggplot2)

base = ggplot(data, aes(x=var))
hist = base + geom_histogram(fill="green", 
                             color='grey',
                          aes(y=..density..))


histAndNormal = hist + stat_function(fun=dnorm,
                                     color="red",
                                     args=list(mean=mnVar,sd=sdVar))

histAndNormal


mdVar=median(var)

histAndNormal + geom_vline(aes(xintercept = mnVar,colour="mean"),
                           show.legend  = TRUE) + 
                geom_vline(aes(xintercept = mdVar,colour="median"),
                           show.legend  = TRUE) + 
                scale_color_manual(name = "centrals", 
                                   values = c(median = "blue", mean = "red"))
```

distribution of other variables

distrubution of gender of head of household
```{r}
table(alldata$hh_b02)
```
much more men than women

recode variables 1=male 2=female
keep numerical and change to 0 and 1 so can get proportions
```{r}
alldata$gender[alldata$hh_b02==1] <- "male"
alldata$gender[alldata$hh_b02==2] <- "female"

alldata$hh_b02[alldata$hh_b02==1] <- 0
alldata$hh_b02[alldata$hh_b02==2] <- 1

```


```{r}
tapply(alldata$expmR, list(hh_b02=alldata$hh_b02,hh_c07=alldata$hh_c07), median)
```


```{r}
library(ggplot2) 

indiv = alldata[alldata$indidy4==1,]
p <- ggplot(indiv, aes(gender, expmR))

p + geom_boxplot() + labs(title="Are the mean consumption different between male headed households and female headed households?")

```

```{r}
seMean = function(x) sd(x)/sqrt(length(x))

# Means by group
means=aggregate(list(mean=alldata$expmR),
          list(gender=alldata$hh_b02),mean)

# se by group: (I am using my fucntion here)
sems=aggregate(list(seMean=alldata$expmR),
          list(gender=alldata$hh_b02),seMean)

# a new data frame:
data=merge(means,sems) # party is the assumed 'key'

# adding confidence interval limits:
data$lower=data$mean-2*data$seMean
data$upper=data$mean+2*data$seMean


data
```

```{r}
meanPlot = ggplot(data, aes(y=mean, x=gender)) + geom_point()

# this adds the error bars:
(errorPlot = meanPlot + geom_errorbar(aes(ymin = lower, ymax = upper)))
```

recode education

```{r}
alldata$educ[alldata$hh_c03==2] <- 0
alldata$educ[alldata$hh_c07<17] <- 1
alldata$educ[alldata$hh_c07==17] <-2  #typical primary education is 7 years
alldata$educ[alldata$hh_c07==18] <-2
alldata$educ[alldata$hh_c07>=21 & alldata$hh_c07<25] <- 3
alldata$educ[alldata$hh_c07==24] <-4
alldata$educ[alldata$hh_c07>=31 & alldata$hh_c07<34] <- 5
alldata$educ[alldata$hh_c07==32] <-6
alldata$educ[alldata$hh_c07==34] <-6
alldata$educ[alldata$hh_c07>34] <- 7

```

recode education level

```{r}
alldata$educ1[alldata$educ==0] <- "no education"
alldata$educ1[alldata$educ==1] <- "some primary"
alldata$educ1[alldata$educ==2] <- "completed primary"
alldata$educ1[alldata$educ==3] <- "some secondary ordinary level"
alldata$educ1[alldata$educ==4] <- "completed secondary ordinary level"
alldata$educ1[alldata$educ==5] <- "some secondary advanced level"
alldata$educ1[alldata$educ==6] <- "completed secondary advanced level"
alldata$educ1[alldata$educ==7] <- "some higher education"

```


```{r}
table(alldata$educ1)
```


```{r}
seMean = function(x) sd(x)/sqrt(length(x))

# Means by group
means=aggregate(list(mean=alldata$expmR),
          list(educ=alldata$educ1),mean)

# se by group: (I am using my fucntion here)
sems=aggregate(list(seMean=alldata$expmR),
          list(educ=alldata$educ1),seMean)

# a new data frame:
data=merge(means,sems) # party is the assumed 'key'

# adding confidence interval limits:
data$lower=data$mean-2*data$seMean
data$upper=data$mean+2*data$seMean


data
```

```{r}

meanPlot = ggplot(data, aes(y=mean, x=educ)) + geom_point()

# this adds the error bars:
(errorPlot = meanPlot + geom_errorbar(aes(ymin = lower, ymax = upper)))
```



```{r}

indiv = alldata[alldata$indidy4==1,]
p <- ggplot(indiv, aes(educ1, expmR))

p + geom_boxplot() + labs(title="Is mean consumption different between education level of head of household?")

```



